---
title: "Practical J: Solutions"
subtitle: "Model estimation"
author: "Maarten Cruyff"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

------------------------------------------------------------------------



```{r message=FALSE, warning=FALSE}
library(MASS)
library(ggplot2)
library(dplyr)
library(gridExtra)
```


# The Diamonds data

In this exercise,  predict the price of diamonds from 
a continuous and categorical variable. The data set is `diamonds` 
of the package `ggplot2`, The data set is very large, so we will work with
a random sample of 500 cases.

First have a look at the structure of the data set with: `str()`. What do you conclude? Which data types do you see?
```{r}
str(diamonds)
```

a.  Set the seed to 100 and---with the help of pipe operators---use the
    `slice_sample()` function to take a random sample form the original data set. 
    Call the new data set: `my_diamonds`,
    Change the class of the variables `cut`, `color` and `clarity` from `ordered factor` to `factor`. 
    Look again at the results with `str()` and `summary()`.


```{r}
set.seed(100)
my_diamonds <- dplyr::slice_sample(diamonds, n = 500, replace = TRUE) %>% 
  dplyr::mutate(cut = factor(cut, ordered = FALSE),
         color   = factor(color, ordered = FALSE),
         clarity = factor(clarity, ordered = FALSE))

str(my_diamonds)
```
```{r}
summary(my_diamonds)
```
## Presentation ready summary tables
The package [`gtsummary`](https://www.danieldsjoberg.com/gtsummary/) provides a flexible way to create presentation-ready summary tables. Have a look at the webpage of this package to learn more about the possibilities.
Install the package `gtsummary` and load it. 

Make a `gtsummary` table of the `my_diamonds` data with the following code: `my_diamonds %>% tbl_summary()`

```{r message=FALSE, warning=FALSE}
# install.packages("gtsummary")
require(gtsummary)

my_diamonds %>% tbl_summary()
```

Make a scatterplot with `ggplot` with the following aesthetics: x-axis = `carat`; y-axis = `price`; color =
`depth` and add a meaningful main title to the plot, as well as axis lables. What do you conclude?


```{r}

p <- ggplot(my_diamonds, aes(x=carat, y=price, color=depth)) +
      geom_point() +
      labs(x = "carat", y = "price", title = "Price of diamonds related to their carat and depth values") 
p
```

------------------------------------------------------------------------

## Numeric predictors

We start with a simple regression predicting the `price` of diamonds from the numeric
predictor `carat`.

a.  Display a scatter plot of `price` against `carat`, and add a linear
    regression line (set the size of the points to 0.5).

```{r}
ggplot(my_diamonds, aes(x = carat, y = price)) +
  geom_point(size = 0.5) +
  geom_smooth(method = "lm", se = FALSE) 
```

b.  Describe the effect of `carat` on `price`.

   - `carat` is positively associated with `price`.


c.  Display and interpret the summary of the model.

```{r}
fit <- lm(price ~ carat, my_diamonds)
fit %>% summary()
```


d.  Test the significance of the model against the intercept-only model.

```{r}
anova(lm(price ~ 1, my_diamonds), fit)
```

e.  Display and interpret a the residual plots of the model.

```{r fig.asp = 1}
par(mfrow = c(2, 2))
plot(fit)
```

## Categorical predictor

We now predict the `price` of diamonds from the categorical
predictors `cut`. The variable `cut` in this data set 
has class `ordered` (the class for
ordinal variables), but, for this practical, we want it to have class
`factor` (the class for categorical variables).

a.  Display boxplots of `price` for the different
    levels of `cut` with `ggplot`. Interpret the results.

```{r}
my_diamonds %>% 
  ggplot(aes(price, cut)) +
  geom_boxplot()

```

b.  Display and interpret the summary of the model with `cut` predicting
    `price`.

```{r}
fit <- lm(price ~ cut, my_diamonds)
fit %>% summary()
```

c.  What is the reference category of `cut`?

   - Fair


d.  Which category has the highest predicted price?

   - Fair


e.  Display and interpret the residual plots of the model.

```{r fig.asp = 1}
par(mfrow = c(2, 2))
plot(fit)
```


## Multiple regression

a.  Fit the main effects model with `carat` and `cut` as predictors of `price`, and display and interpret the summary of the regression results. 


```{r}
fit_main <- lm(price ~ carat + cut, my_diamonds) 
summary(fit_main)
```
b. Make a table of the output of the regression model with `tbl_regression()` (part of the `gtsummary` package) using the following code:
```{r}
t1 <-
  lm(price ~ carat + cut, my_diamonds) %>% 
  tbl_regression()
t1
```


b.  Qualitatively compare the size and significance of the parameter
    estimates from this model to those from the simple regression
    models.


   - The effects are somewhat smaller in the multiple regression model and the 
     significance levels are somewhat lower. The `depth` effect was significant 
     in the simple regression model and not in the multiple regression model.



a.  Fit a the same model with the interaction between `carat` and `cut` included, and display and interpret the model
    summary.

```{r}
fit_interact <- lm(price ~ carat * cut, my_diamonds) 
summary(fit_interact)
```

```{r}
t2 <-
  lm(price ~ carat * cut, my_diamonds) %>% 
  tbl_regression()
t2
```

b.  What do you notice about the significance of the effects?

   - Only the effect of `carat` is significant.

c. Compare the fit of the model with main effects and the model with interaction effects with the `anova` function.
```{r}
anova(fit_main, fit_interact)
```
Make a presentation-ready regression table of both models: the main effect model and the interaction effect model with the following code:

```{r}
# presentation ready regression table

tbl_merge(
  list(t1, t2),
  tab_spanner = c("**Main effects model**", "**Interaction model**")
)
```


d.  Display a scatter plot with `price` on the x-axis, `carat` on the y-axis, with
different colors for the points and linear regression lines for the levels of `cut`.
Suppress the confidence bands. What does this plot tell you about the main effects 
and the interaction effect of `carat` and `cut`?

```{r}
ggplot(my_diamonds, aes(carat, price, col = cut))+
  geom_point() +
  geom_smooth(method = "lm", se = F)
```

   - The regression lines are separated, so there are some main effects.
   - They are almost parallel, so no interactions.


End of practical
