---
title: "Practical G: Template"
author: "Gerko Vink"
date: "Statistical Programming with R"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
printAnswers <- TRUE
```
---

We use the following packages in this Practical:
```{r warning=FALSE, message=FALSE, error=FALSE}
library(MASS)
library(plyr)
library(dplyr)
library(magrittr)
library(ggplot2)
```

If you need package `plyr` and package `dplyr`, always load package `plyr` first!
If you load `dplyr` before `plyr`, `plyr` will produce a warning. 

---

#### Exercises

---

In this exercise we will again be using random number generators. When using random numbers, it is wise to always fix the random seed (the starting point of the number generation). This ensures that in the future we can exactly reproduce the chain of executions that led to the randomly generated results. 

Start by setting a random seed. If you follow my random seed and reproduce the code below in exactly the same order, you will get the same results. If you do not follow the exact ordering (i.e. if you skip or rerun a question, or have different code), your results may be different due to random sampling. This is not a bad thing! It is just a result of the method and should be that way.

```{r}
#set random seed, make things reproducible
set.seed(123)
```

---

1. **A group of experimenters have 4 experimental conditions they want to run. In each of these four conditions, there are seven manipulations that should be in a random order. Design the experiment such that for every condition, the seven manipulations are randomly ordered.** 

```{r eval=printAnswers, echo=printAnswers}
# Make an empty matrix. 

# add your code here

# Name the columns and rows. 
colnames(design) <- paste("condition", 1:4, sep=" ")
rownames(design) <- paste("manipulation",  1:7, sep=" ")

# Put a random ordering in each column.
# add your code here

design
```

---

2. **Generate a vector of 100 random standard normal numbers.**

```{r eval=printAnswers, echo=printAnswers}
# Get random numbers.
y <- rnorm(100)
```

---

3. **Compute the mean and standard deviation of the vector from question 2.**

```{r eval=printAnswers, echo=printAnswers}
# add your code here
```

---

4. **Generate a vector of 100 random standard normal numbers (like in Exercise 2) 25 times, and each time store means in object `av`. Compute the standard deviation of `av`.**

```{r eval=printAnswers, echo=printAnswers}
# Repeat the sampling of numbers 25 times, each time getting the mean.
# add your code here

# Standard error of sample mean.
# add your code here
```

or alternatively with function `rlply()` from package `plyr`:
```{r}
samples <- rlply(.n = 25, rnorm(100, mean = 0, sd = 1))
```

Function `plyr::rlply()` evaluates an expression $n$ times and combines the 
results in a list. So the call `rnorm(5000, mean = 0, sd = 1)` is evaluated
`.n = 25` times. 

We can then call the following `sapply()` statement to create object `av`:
```{r}
av <- sapply(samples, mean)
sd(av) # Standard error of sample mean.
```

Function `sapply()` evalutes the expression `mean()` over each listed element in 
`samples`. The result is a vector of length `100` with the means of the 100 samples
from the `samples` object.

---

5. **Create a function that automatically returns a vector like `av`.**

```{r eval=printAnswers, echo=printAnswers}

    # add your code here
 
```

or, with `rlply()`:

```{r eval=printAnswers, echo=printAnswers}

  # add your code here



```

---

6. **Add the option to this function to print a density plot. Set it to TRUE by default. **

```{r eval=printAnswers, echo=printAnswers}
# Create a function for sd of sample means

```

The above function does two things:

1. It returns a vector of 25 means by default (the default argument for reps is `reps=25`)
2. It plots a densityplot of those 25 means, if wanted. By default it does this because the argument for plotting is `plotDens=TRUE`. 

Let's go through the function. 

- `av <- numeric(reps)`. This line creates a vector with zeros that is long as the number of `reps`. 
- `for(i in 1:reps) {`. This line starts the for loop and dictates that we are goint to repeat the code within the loop 1 through 25 - so 25 times in total. 
- `y <- rnorm(n)`. This samples `n` values from the standard normal distribution. Remember that the default argument for `n` is `n=100`. So, by default it draws `100` values. These values are stored in object `y`, such that we can use it later on in the function. 
- `av[i] <- mean(y)`. Now we calculate the mean on the current (*i*th) sample and store that in the vector `av` as the *i*th element. Over all the for-loops we will replace each element in `av` with the mean of the respective simulated sample. 
- `if (plotDens) {` evaluates whether to execute the code within the `{` and `}` that define the if-statement. This code is designed to print a plot. The default for `plotDens` is `plotDens=TRUE`, so by default the if-statement will print the plot. `if` looks for `TRUE` or `FALSE` and only executes its code when it finds `TRUE`. Because `plotDens` is `TRUE`, we do not have to say `if (plotDens = TRUE)` and `if(plotDens)` suffices. 
- `plot(density(av), main = "Sampling distribution of the mean.")` plots a densityplot with main title 'Sampling distribution of the mean.'. 

or, with `rlply()` and a pipe:

```{r eval=printAnswers, echo=printAnswers}
# Create a function for sd of sample means
# add your code here
```

---

In the next codeblock, we first set the graphical parameters to display a 2 by 2 matrix of plots. Then we run the function `mean.av()` 4 times; each time generating a plot. After running the function four times, we will have filled the 2 by 2 plot raster.  We do not have to specify the arguments of this function because all arguments are set by default. We end by resetting the graphical parameters to its original state by stating `par(mfrow = c(1, 1))`, such that when we plot a new graph, it will be displayed as a single graph and not in a raster of 4 plots. 
```{r}
# add your code here
```

and we return the graphing parameter to its previous state of a single plot (i.e. 1 row and 1 column). 
```{r}
# add your code here
```

---

7. **Generate a random sample of size 20 from a normal population with mean 100 and standard deviation 10. **

```{r eval=printAnswers, echo=printAnswers}
# add your code here
```

---

8. **Use `mfrow` to set up the layout for a 3 by 4 array of plots. In the top 4 panels, show normal probability plots ('QQ-plots') for 4 separate "random" samples of size 10, all drawn from a normal distribution. In the middle 4 panels, display plots for samples of size 100. In the bottom 4 panels, display plots for samples of size 1000. Comment on how the appearance of the plots changes as the sample size changes.**

```{r eval=printAnswers, echo=printAnswers}
# add your code here
```

---

9. **Repeat exercise 8, but use `runif` instead of `rnorm`.**

```{r eval=printAnswers, echo=printAnswers}
# add your code here
```

---

10. **Use the function `rexp()` to simulate 100 exponential random numbers with rate 0.2. Do the following on the simulated random numbers **

- Obtain a density plot for the observations. 
- Find the sample mean of the observations. 
- Compare with the population mean (the mean for an exponential population is 1/rate).

```{r eval=printAnswers, echo=printAnswers}
# add your code here
```

---

11. **Fit the following linear models on the anscombe data:**

- `y1` predicted by `x1` - stored in object `fit1`
- `y2` predicted by `x2` - stored in object `fit2`
- `y3` predicted by `x3` - stored in object `fit3`
- `y4` predicted by `x4` - stored in object `fit4`

```{r}
# add your code here
```

---

12. **`Create a data frame from the coefficients of the 4 fitted objects from Exercise 11**
```{r}
# add your code here
```

These estimates are very similar. 

---

13. **Plot the four fitted models from Exercise 11 in a single plotting window. Make the points in the plots `blue`, `gray`, `orange` and `purple`, respectively. **
```{r}
# add your code here
```

---

14. **Now plot all four fitted models from Exercise 11 in a plotting window with 2 rows and 2 columns.**
```{r}
# add your code here
```

---

End of practical. 
