---
title: "Practical F: template"
author: "Gerko Vink"
date: "Statistical Programming with R"
output: html_document
---

---

#### Exercises

---

The following packages are required for this practical:
```{r, message=FALSE}
library(dplyr)
library(magrittr)
library(mice)
```
and if you'd like the same results as I have obtained, you can fix the random seed
```{r}
set.seed(123)
```

---

1. **Use a pipe to do the following:**

- draw 1000 values from a normal distribution with `mean = 5` and `sd = 1` - $N(5, 1)$, 
- create a matrix where the first 500 values are the first column and the second 500 values are the second column **
- make a scatterplot of these two columns
```{r}

```


---

2. **Use a pipe to calculate the correlation matrix on the `anscombe` data set**

```{r}

```

---

3. **Now use a pipe to calculate the correlation for the pair (`x4`, `y4`) on the `anscombe` data set**

Using the standard `%>%` pipe:
```{r}

```
Alternatively, we can use the `%$%` pipe from package `magrittr` to make this process much more efficient.
```{r}

```

---

4. **Use a pipe to calculate the correlation between `hgt` and `wgt` in the `boys` data set from package `mice`.**

```{r}

```

---

5. **In the `boys` data set, `hgt` is recorded in centimeters. Use a pipe to transform `hgt` in the `boys` dataset to height in meters and verify the transformation**


```{r}

```

---

End of Practical
 
---

#### Useful References

- [`magrittr`](https://magrittr.tidyverse.org)
- [`R` for Data Science](http://r4ds.had.co.nz) - [Chapter 18 on pipes](http://r4ds.had.co.nz/pipes.html)